\documentclass{beamer}

\usetheme{Madrid}
\usecolortheme{default}

\usepackage[utf8]{inputenc}
\usepackage{braket}
\usepackage{amsmath, amsfonts}

\title[Quantum Error Correction]{Quantum Error Correction and Entanglement Purification}
\author[EE7001 Project]{Tanay Bhat \and Jay Mehta \and Rajwardhan Toraskar}
\institute[IIT Bombay]{Department of Electrical Engineering\\
Indian Institute of Technology Bombay}
\date{}

\begin{document}

%----------------- Title -----------------
\begin{frame}
  \titlepage
\end{frame}

%----------------- Outline -----------------
\begin{frame}{Outline}
  \tableofcontents
\end{frame}

\section{Motivation and Overview}

\begin{frame}{Motivation}
  \begin{itemize}
    \item Quantum information is fragile and susceptible to errors which may arise due to faulty gates, environmental decoherence etc.
    \item Reliable quantum computation and communication require active protection of quantum states which maybe achieved through:
    \begin{itemize}
      \item \textbf{Quantum Error Correction (QEC)}: encode logical qubits into many physical qubits, detect and correct errors.
      \item \textbf{Entanglement Purification}: distill high-fidelity entangled pairs from many noisy copies.
    \end{itemize}
    \item In the project, we shall explore some QEC codes and bipartite purification protocols, and implement them in Qiskit. We shall also establish the equivalence between one-way entanglement purification and quantum error correction codes.
  \end{itemize}
\end{frame}

\section{Quantum Errors}

\begin{frame}{Types of Quantum Errors - Coherent Errors}
  \begin{itemize}
    \item These errors arise from imperfect control of quantum gates, leading to small deviations in the intended operations.
    \item For instance, consider a small rotation $e^{i\epsilon X}$ applied N times to $\ket{0}$.
    \item We have the following expression for the final state:
    \begin{equation}
        \ket{\psi} = \prod_{i = 1}^{N} e^{i\epsilon X} \ket{0} = \cos(N\epsilon)\ket{0} + i\sin(N\epsilon)\ket{1}
    \end{equation}
    \item The probability of measuring $\ket{1}$ is given by:
    \begin{equation}
        P(\ket{1}) = \sin^2(N\epsilon) \approx N^2\epsilon^2 \quad \text{for small } \epsilon.
    \end{equation}
    \item Hence, the error probability grows as $\mathcal{O}(N^2 \epsilon^2)$.
  \end{itemize}
\end{frame}

\begin{frame}{Types of Quantum Errors - Environmental Decoherence}
  \begin{itemize}
    \item Quantum systems can become entangled with their environment, leading to loss of coherence.
    \item For example, consider an environment with states $\ket{e_0}$ and $\ket{e_1}$ such that the environment state flips when the qubit is in state $\ket{1}$:
    \item On applying the operation $HIH$ (ideally identity) on the qubit $\ket{0}\ket{e_0}$, we have:
    \begin{equation}
        \ket{\psi} = HIH \ket{0}\ket{e_0} = \frac{1}{2}(\ket{0} + \ket{1})\ket{e_0} + \frac{1}{2}(\ket{0} - \ket{1})\ket{e_1}
    \end{equation}
    \item Tracing out the environment, the reduced density matrix of the qubit becomes:
    \begin{equation}
        \rho = \frac{1}{2}(\ket{0}\bra{0} + \ket{1}\bra{1})
    \end{equation}
  \end{itemize}
\end{frame}

\begin{frame}{Types of Quantum Errors - Other Imperfections}
  \begin{itemize}
    \item \textbf{Measurement errors:} Faulty measurement apparatus can yield incorrect outcomes. \begin{itemize}
        \item Consider the operators:
        \begin{equation*}
            \begin{aligned}
                F_0 = (1-p)\ket{0}\bra{0} + p\ket{1}\bra{1} \\
                F_1 = p\ket{0}\bra{0} + (1-p)\ket{1}\bra{1}
            \end{aligned}
        \end{equation*}
        \item The final state after measurement will be a superposition state weighted by these probabilities.
    \end{itemize}
    \item \textbf{Qubit loss:} Physical qubits may be lost due to decoherence or operational errors, leading to incomplete information about the quantum state.
    \item \textbf{Initialization errors:} Errors during state preparation can lead to incorrect initial states, affecting subsequent computations.
    \item \textbf{Qubit Leakage:} Qubits may leak out of the computational subspace into higher energy levels, which may introduce unwanted dynamics in a quantum circuit.
  \end{itemize}
\end{frame}

\section{3- and 9-qubit Codes}

\begin{frame}{3-qubit Bit-flip Code}
  \begin{itemize}
    \item Encodes one logical qubit into three physical qubits:
    \[
      \ket{0}_L = \ket{000}, \quad \ket{1}_L = \ket{111}.
    \]
    \item \textbf{Distance} $d = 3 \Rightarrow$ corrects one bit-flip error ($t = \lfloor (d-1)/2 \rfloor = 1$).
    \item Error detection via parity checks using ancilla qubits.
    \item For coherent $X$-rotations on each qubit:
    \[
      F_{\text{unencoded}} \approx 1 - \epsilon^2, \qquad
      F_{\text{encoded}} \approx 1 - \epsilon^6
    \]
    (after successful syndrome measurement and correction).
  \end{itemize}
\end{frame}

\begin{frame}{9-qubit Shor Code}
  \begin{itemize}
    \item Combines 3-qubit bit-flip and 3-qubit phase-flip codes.
    \[
      \ket{0}_L = \frac{1}{2\sqrt{2}}\bigl(\ket{000}+\ket{111}\bigr)^{\otimes 3}
    \]
    \[
      \ket{1}_L = \frac{1}{2\sqrt{2}}\bigl(\ket{000}-\ket{111}\bigr)^{\otimes 3}
    \]
    \item Corrects any single-qubit error (arbitrary Pauli $X,Y,Z$).
    \item Bit-flip correction: apply 3-qubit repetition code within each block.
    \item Phase-flip correction: compare phases between the three blocks.
  \end{itemize}
\end{frame}

\section{Stabilizer Codes}

\begin{frame}{Stabilizer Formalism}
  \begin{itemize}
    \item A state $\ket{\psi}$ is \emph{stabilized} by $K$ if $K\ket{\psi} = \ket{\psi}$.
    \item Stabilizer group $G$: Abelian subgroup of the $n$-qubit Pauli group $P_n$.
    \item Code space: simultaneous $+1$ eigenspace of all generators.
    \item For $n$ physical qubits and $k$ independent generators:
    \[
      \dim(\text{code space}) = 2^{n-k}
    \]
  \end{itemize}
\end{frame}

\begin{frame}{Steane [[7,1,3]] Code}
  \begin{itemize}
    \item CSS stabilizer code with 6 generators:
    \begin{itemize}
      \item 3 $X$-type stabilizers.
      \item 3 $Z$-type stabilizers.
    \end{itemize}
    \item Encodes 1 logical qubit, distance $d=3$ (corrects 1 arbitrary Pauli error).
    \item Logical operators: e.g.\ $Z_L = Z^{\otimes 7}$, $X_L = X_1 X_2 X_3$.
    \item State preparation:
    \begin{itemize}
      \item Start from $\ket{0}^{\otimes 7}$.
      \item Measure $X$-type stabilizers with ancillas.
      \item Apply classically controlled $Z$ corrections to enforce $+1$ eigenvalues.
    \end{itemize}
  \end{itemize}
\end{frame}

\section{Digitization of Quantum Noise}

\begin{frame}{Digitization via Syndrome Measurement}
  \begin{itemize}
    \item Realistic noise is continuous (coherent rotations, Lindblad evolution).
    \item Any CPTP map can be expanded in the Pauli basis:
    \[
      \rho \rightarrow \sum_k A_k \rho A_k^\dagger,\quad A_k = \sum_j \alpha_{k,j} P_j.
    \]
    \item Syndrome measurement projects superpositions of errors onto definite Pauli errors:
    \[
      \rho \rightarrow P_j \rho P_j \quad \text{with probability } |\alpha_j|^2.
    \]
    \item QEC effectively reduces continuous errors to \textbf{stochastic Pauli faults} with some error rate $p$.
  \end{itemize}
\end{frame}

\section{Entanglement Purification}

\begin{frame}{Werner States}
  \begin{itemize}
    \item Start from general mixed state in Bell basis:
    \[
      \rho'_{AB} = \sum_{k_1,k_2,j_1,j_2} \lambda_{k_1k_2j_1j_2}
      \ket{\phi_{k_1k_2}}\bra{\phi_{j_1j_2}}.
    \]
    \item Random bilateral operations can be used to remove off-diagonal terms
    without changing fidelity with $\ket{\phi_{00}}$.
    \item Resulting Werner state:
    \[
      \rho_W(x) = x \ket{\phi_{00}}\bra{\phi_{00}} + \frac{1-x}{4} I_4.
    \]
    \item Any purification protocol that works for Werner states works for all
    states with the same fidelity.
  \end{itemize}
\end{frame}

\begin{frame}{BBPSSW Protocol}
  \begin{itemize}
    \item Input: two copies of a Werner state with fidelity $F > 1/2$.
    \item Steps:
    \begin{enumerate}
      \item Depolarize (twirl) to Werner form.
      \item Apply bilateral CNOT: $A_1 \rightarrow A_2$, $B_1 \rightarrow B_2$.
      \item Measure target pair; keep source pair only for certain outcomes.
    \end{enumerate}
    \item Updated fidelity (conditional on success):
    \[
      F' = \frac{F^2 + \bigl(\frac{1-F}{3}\bigr)^2}
                 {F^2 + \frac{2F(1-F)}{3} + 5\bigl(\frac{1-F}{3}\bigr)^2}.
    \]
    \item Fidelity increases with each successful round; yield decreases due to discarding pairs.
  \end{itemize}
\end{frame}

\begin{frame}{DEJMPS Protocol}
  \begin{itemize}
    \item Designed for Bell-diagonal states; more efficient than BBPSSW.
    \item Key ideas:
    \begin{itemize}
      \item Local basis rotations convert phase errors into bit errors.
      \item Bilateral CNOT and post-selection on matching measurement outcomes.
    \end{itemize}
    \item Coefficients in Bell basis $\{\lambda_{00},\lambda_{01},\lambda_{10},\lambda_{11}\}$ update as:
    \[
      \lambda'_{00} = \frac{\lambda_{00}^2 + \lambda_{11}^2}{N},\quad
      \lambda'_{10} = \frac{2\lambda_{00}\lambda_{11}}{N},
    \]
    \[
      \lambda'_{01} = \frac{\lambda_{01}^2 + \lambda_{10}^2}{N},\quad
      \lambda'_{11} = \frac{2\lambda_{01}\lambda_{10}}{N},
    \]
    \[
      N = (\lambda_{00} + \lambda_{11})^2 + (\lambda_{01} + \lambda_{10})^2.
    \]
    \item Converges to $\ket{\Phi^+}$ for initial fidelity $F = \lambda_{00} > 1/2$.
  \end{itemize}
\end{frame}

\section{One-way Protocols and QECC}

\begin{frame}{One-way Hashing Protocol}
  \begin{itemize}
    \item Alice and Bob share $n$ noisy Bell pairs (Bell-diagonal state $\rho$).
    \item Represent each pair by 2-bit error label $(i,j)$ (phase and amplitude errors).
    \item Repeatedly:
    \begin{itemize}
      \item Choose random parity vector $s$.
      \item Use local unitaries + bilateral CNOT to map $s \cdot x$ into one pair.
      \item Measure and discard that pair; record the parity.
    \end{itemize}
    \item After sacrificing $\approx nS(\rho)$ pairs, remaining $m \approx n(1-S(\rho))$
    pairs can be corrected.
    \item \textbf{Asymptotic yield}: $m/n \approx 1 - S(\rho)$.
  \end{itemize}
\end{frame}

\begin{frame}{Equivalence of 1-EPP and QECC}
  \begin{itemize}
    \item Let $Q(\chi)$ be the quantum capacity of a channel $\chi$.
    \item Let $D(M)$ be the one-way distillable entanglement of a mixed state $M$.
    \item Using teleportation with noisy entangled pairs:
    \begin{itemize}
      \item Any QECC of rate $R$ yields a 1-EPP with yield $\ge R$:
      \[
        Q(\chi(M)) \le D(M).
      \]
      \item Any 1-EPP with yield $R$ yields a QECC of rate $\ge R$:
      \[
        D(M) \le Q(\chi(M)).
      \]
    \end{itemize}
    \item Therefore: \textbf{one-way entanglement purification and QECC are equivalent}
    in terms of achievable rates.
  \end{itemize}
\end{frame}

\begin{frame}{Equivalence Proof: $Q(\chi(M)) \le D(M)$ }
    Consider a quantum teleportation setup where Alice and Bob share multiple copies of a mixed entangled state $M$ instead of perfect Bell pairs. We construct a purification protocol as follows:
  \begin{itemize}
    \item Alice prepares to send $n$ qubits by preparing $m$ Bell pairs and $n-m$ ancilla qubits in state $\ket{0}$. Encode one half of each Bell pair and the ancillas using a QECC of rate $R = m/n$.
    \item Alice sends her encoded qubits through the channel $\chi(M)$ to Bob.
    \item Bob receives the qubits and performs error correction using the QECC and recovers the $m$ halves of the Bell pairs.
  \end{itemize}
  Hence $n$ copies of the mixed state $M$ were used to get $m$ high-fidelity Bell pairs, yielding a distillation rate of $R = m/n$. The yield of such a 1-EPP is at least as large as the rate of the QECC used, leading to: 
    \[
        Q(\chi(M)) \le D(M).
    \]
\end{frame}

\begin{frame}{Equivalence Proof: $Q(\chi(M)) \ge D(M)$}
    Consider a one-way entanglement purification protocol (1-EPP) that distills $m$ high-fidelity Bell pairs from $n$ copies of a mixed state $M$. We construct a QECC as follows:
  \begin{itemize}
    \item Alice and Bob use the 1-EPP to distill $m$ high-fidelity Bell pairs from $n$ copies of $M$.
    \item Alices prepares to send $m$ qubits of quantum information. She uses the distilled Bell pairs to teleport these qubits to Bob.
  \end{itemize}
  Hence we are able to send $m$ qubits of quantum information without error using $n$ uses of the channel $\chi(M)$, yielding a quantum communication rate of $R = m/n$. The rate of such a QECC is at least as large as the yield of the 1-EPP used, leading to:
    \[
        D(M) \le Q(\chi(M)).
    \]
\end{frame}

\section{Simulations and Results}

\begin{frame}{Qiskit Implementations}
  \begin{itemize}
    \item Implemented in Qiskit:
    \begin{itemize}
      \item 3-qubit bit-flip code (X error and coherent rotation suppression).
      \item 9-qubit Shor code for correcting a $Z$ error.
      \item Steane [[7,1,3]] code state preparation for $\ket{0}_L$ and $\ket{1}_L$.
      \item BBPSSW and DEJMPS entanglement purification circuits.
    \end{itemize}
    \item Results:
    \begin{itemize}
      \item Histograms show deterministic recovery of logical states in 3- and 9-qubit codes.
      \item Numerical fidelity matches theoretical predictions ($1-\epsilon^6$ vs $1-\epsilon^2$).
      \item Purification protocols increase fidelity for initial $F>0.5$, with non-unit success probability.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Conclusion}
  \begin{itemize}
    \item Demonstrated how:
    \begin{itemize}
      \item Simple repetition and Shor codes protect against bit-flip and phase-flip errors.
      \item Stabilizer formalism generalizes QEC and simplifies encoding/decoding.
      \item Syndrome measurements digitize continuous noise into discrete Pauli errors.
      \item Entanglement purification (BBPSSW, DEJMPS, hashing) enhances entanglement fidelity.
    \end{itemize}
    \item Established conceptual link and rate-equivalence between 1-EPP and QECC.
    \item Simulations validate theoretical performance of the studied codes and protocols.
  \end{itemize}
\end{frame}

\end{document}
